---
title: "Codon Usage Bias Analysis Project with AnaCoDa"
author: "Mia Papazian"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
knitr::opts_knit$set(root.dir = "~/repositories/mias_intragenomic_variation_mutation_bias")
library(AnaCoDa)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
```

# 1. data preparation

## load genome
```{r genome, eval=FALSE}
genome <- initializeGenomeObject(file = "fasta/revisit_cds_data/candida_tenuis.max.cds")
```

## initialize parameters
```{r parameter, eval=FALSE}
parameter <- initializeParameterObject(
  genome = genome, 
  sphi = 1, 
  num.mixtures = 1, 
  gene.assignment = rep(1, length(genome))
)
```

## initialize MCMC and model
```{r mcmc_model, eval=FALSE}
mcmc <- initializeMCMCObject(samples = 100, thinning = 10, adaptive.width = 50)
model <- initializeModelObject(parameter = parameter, model = "ROC")
```

# 2. run MCMC

```{r run_mcmc, eval=FALSE}
runMCMC(mcmc = mcmc, genome = genome, model = model)
```

# 3. save/load results

```{r save_results, eval=FALSE}
species <- "candida_tenuis"
dir.create("Results", showWarnings = FALSE)

mcmc_file <- file.path("Results", paste0(species, "_mcmc.Rda"))
writeMCMCObject(mcmc = mcmc, file = mcmc_file)

param_file <- file.path("Results", paste0(species, "_parameter.Rda"))
writeParameterObject(parameter = parameter, file = param_file)
```

```{r load_results}
#working on this section
```

# 4. trace analysis

## extract trace object
```{r trace}
trace <- parameter$getTraceObject()
plot(x = trace, what = "Expression", mixture = 1, geneIndex = 1)
```

## synthesis rate trace (φ)
```{r synthesis_trace}
synth_trace_list <- trace$getSynthesisRateTrace()[[1]]

trace_to_tibble <- function(list_trace, id_names) {
  matrix <- do.call(rbind, list_trace)
  sample_n <- 1:ncol(matrix) - 1
  colnames(matrix) <- sample_n
  tibble <- as_tibble(matrix) %>% bind_cols(tibble(id_names), .)
  return(tibble)
}

synth_trace <- trace_to_tibble(synth_trace_list, id_names = getNames(genome))
head(synth_trace)
dim(synth_trace)
```

# 5. codon-specific parameters

## extract CSP estimates
```{r csp_estimates}
csp_est <- getCSPEstimates(parameter = parameter, mixture = 1, samples = 100)
csp_codons <- csp_est$Mutation %>% filter(Mean != 0) %>% select("Codon")
str(csp_codons)
print(csp_codons)
```

## extract codon-specific traces
```{r csp_trace}
csp_cat <- c("Mutation", "Selection")
csp_wide_list <- list()

for (i in 1:2) {
  param = csp_cat[i]
  trace_list <- trace$getCodonSpecificParameterTrace((i - 1))[[1]]
  csp_wide_list[[i]] <- trace_to_tibble(trace_list, csp_codons) %>%
    rowwise() %>%
    mutate(AA = codonToAA(Codon), .before = "Codon")
}

names(csp_wide_list) <- csp_cat
```

## pivot traces to long format
```{r csp_long}
csp_long_list <- list()

for (i in 1:2) {
  param = csp_cat[i]
  csp_long_list[[i]] <- pivot_longer(csp_wide_list[[i]],
                                     cols = !c("AA", "Codon"),
                                     values_to = param,
                                     names_to = "Sample") %>%
    mutate(Sample = as.numeric(Sample))
}

names(csp_long_list) <- csp_cat

csp_long <- left_join(csp_long_list$Mutation,
                      csp_long_list$Selection,
                      by = join_by(AA, Codon, Sample))
head(csp_long)
```

#CSP parameter comparison
```{r csp_comparison}
#get anacoda’s estimates
csp_estimates <- getCSPEstimates(parameter, mixture = 1, samples = 100)

#combine mutation and selection estimates
csp_summary_ac <- full_join(csp_estimates$Mutation %>% rename(Mutation_mean = Mean),
                            csp_estimates$Selection %>% rename(Selection_mean = Mean),
                            by = c("AA", "Codon"))

#now calculate my means from the trace tibble
csp_summary_tibble <- csp_long %>%
  group_by(AA, Codon) %>%
  summarize(Mutation_calc = mean(Mutation), Selection_calc = mean(Selection), .groups = "drop")

# join the two
csp_comparison <- inner_join(csp_summary_ac, csp_summary_tibble, by = c("AA", "Codon"))

# plot: Mutation
ggplot(csp_comparison, aes(x = Mutation_mean, y = Mutation_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Mutation Parameter: AnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")

# plot: Selection
ggplot(csp_comparison, aes(x = Selection_mean, y = Selection_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Selection Parameter: AnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")

```
to validate the accuracy of my trace extraction, i compared codon-specific parameter estimates derived manually from the MCMC traces (csp_long) against AnaCoDa’s built-in summary statistics from getCSPEstimates(). the first figure displays a scatterplot comparing mutation parameters across codons, while the second shows the same comparison for selection parameters.

# 6. double checking codons
```{r correct_codons}
#filter out reference codons which have a mean of 0
csp_codons <- csp_est$Mutation %>%
  filter(Mean !=0) %>%
  select("Codon")
print(csp_codons)

# convert to character vector
csp_codon_list <- csp_codons$Codon
trace_codon_list <- unique(csp_long$Codon)
```

```{r correct_codons_check}
# check: are they exactly the same set?
setequal(csp_codon_list, trace_codon_list)
```

```{r codon_counts}
#getting codon names 
codon_names <- csp_codons
#using csp_codons

#full 64-codon count
full_codon_counts <- getCodonCounts(genome)

#getting codon counts and gene ids
codon_names_vec <- codon_names$Codon
codon_counts <- full_codon_counts[, codon_names_vec] #filtered 40
str(codon_counts)
```

# 7. double checking objects 

```{r double_check_str_synth, echo=TRUE, message=TRUE, warning=TRUE}
# structure and data previews for key objects
str(synth_trace)
```

```{r double_check_idnames_synth, echo=TRUE, message=TRUE, warning=TRUE}
head(synth_trace$id_names)
```

```{r double_check_colnames_synth, echo=TRUE, message=TRUE, warning=TRUE}
colnames(synth_trace)[2:6]
```

```{r double_check_cc, echo=TRUE, message=TRUE, warning=TRUE}
str(codon_counts)
```

```{r double_check_rownames_cc, echo=TRUE, message=TRUE, warning=TRUE}
rownames(codon_counts)[1:3]
```

```{r double_check_colnames_cc, echo=TRUE, message=TRUE, warning=TRUE}
colnames(codon_counts)[1:6]
```

```{r double_check_length_genenames, echo=TRUE, message=TRUE, warning=TRUE}
length(gene_names)
```

```{r double_check_genenames, echo=TRUE, message=TRUE, warning=TRUE}
head(gene_names)
```


```{r double_check_str_csplong, echo=TRUE, message=TRUE, warning=TRUE}
str(csp_long)
```


```{r double_check_csplong, echo=TRUE, message=TRUE, warning=TRUE}
head(csp_long)
```


```{r double_check_first_sample_ids, echo=TRUE, message=TRUE, warning=TRUE}
unique(csp_long$Sample)[1:5]
```

# 8. log likelihood loop 
```{r random_subset}
#retrieve random subset of 300 genes 
#convert csp_codons tibble to character vector
csp_codon_list <- csp_codons$Codon

#filter: keep only genes that have non-zero counts for at least one modeled codon
valid_gene_mask <- apply(codon_counts[, csp_codon_list], 1, function(x) any(x > 0))
valid_gene_names <- rownames(codon_counts)[valid_gene_mask]

#sample: select 300 random valid genes
set.seed(42)  # reproducibility
subset_gene_names <- sample(valid_gene_names, size = 300)
head(subset_gene_names)

#full subset of 4945 genes
# gene_names
# str(gene_names)
```

```{r llik_loop, message=TRUE, warning=TRUE}
loglik_list <- list()
row_index <- 1

for (gene in subset_gene_names) {
  # get codon counts for this gene
  counts <- as.numeric(codon_counts[gene, , drop = TRUE])
  gene_codons <- colnames(codon_counts)
  names(counts) <- gene_codons
  
  for (s in 0:100) {
    # get φ (synthesis rate) for this gene and sample
    phi <- synth_trace %>%
      filter(id_names == gene) %>%
      pull(as.character(s)) %>%
      as.numeric()
    
    # get mutation and selection params for this gene/sample
    dM <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Mutation)
    
    dE <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Selection)
    
    # compute log-likelihood
    ll_result <- LLikGene(gene, dM, dE, phi, counts)
    
    # store result
    loglik_list[[row_index]] <- data.frame(
      Gene = gene,
      Sample = s,
      LogLikelihood = ll_result$LogLikelihood
    )
    
    row_index <- row_index + 1
  }
}

# combine results into a tibble
loglik_df <- dplyr::bind_rows(loglik_list)

print(loglik_df)
```

# 9. for each sample, sum log-likelihoods across all genes 
```{r sum_llik, message=TRUE, warning=TRUE}
loglik_summary <- loglik_df %>%
  group_by(Sample) %>%
  summarise(LL_sum = sum(LogLikelihood), .groups = "drop") 
#get rid of grouping to avoid issues

head(loglik_summary)
#now i have one total log-likelihood value per sample
```

# 10. get anacodas loglikelihood trace 
```{r anacoda_llik_trace, message=TRUE, warning=TRUE}
# extract anacodas log likelihood trace from the MCMC object
logLikeTrace <- mcmc$getLogLikelihoodTrace()

#create tibble for comparison
logLikeTrace_df <- tibble(
  Sample = 1:100,
  AnaCoDa_LL = logLikeTrace[-1] # drop the first zero
)
#check
head(logLikeTrace_df)
```

# 11. plotting my ll vs. anacoda ll
```{r myll_vs_anacodall, message=TRUE, warning=TRUE}
# load full per-gene log-likelihood dataframe
loglik_df_full <- readRDS("loglik_df_fullgenome.rds")

# 1. summarize total log-likelihood per sample from full genome data
# (this happens after adjustment now — see step 7)

# 2. build AnaCoDa LL tibble (drop burn-in)
logLikeTrace_df <- tibble(
  Sample = 1:100,
  AnaCoDa_LL = logLikeTrace[-1]
)

# 3. compute multinomial constant per gene (only depends on codon counts)
log_multinomial_constant <- function(counts) {
  total <- sum(counts)
  if (total == 0) return(0)
  lgamma(total + 1) - sum(lgamma(counts + 1))
}

gene_constants <- tibble(
  Gene = rownames(codon_counts),
  Constant = apply(codon_counts, 1, log_multinomial_constant)
)

# 4. join constants into your existing log-likelihoods and adjust
loglik_df_adjusted <- loglik_df_full %>%
  left_join(gene_constants, by = "Gene") %>%
  mutate(AdjustedLogLikelihood = LogLikelihood + Constant)

# 5. re-summarize per sample using adjusted values
loglik_full_summary <- loglik_df_adjusted %>%
  group_by(Sample) %>%
  summarise(LL_sum = sum(AdjustedLogLikelihood), .groups = "drop")

# 6. compare to AnaCoDa's LL trace
comparison <- left_join(logLikeTrace_df, loglik_full_summary, by = "Sample")

# 7. plot adjusted LLs vs AnaCoDa LLs
ggplot(comparison, aes(x = AnaCoDa_LL, y = LL_sum)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Adjusted Custom Log-Likelihood vs AnaCoDa Log-Likelihood",
    subtitle = "After Adding Multinomial Constants (Per Gene)",
    x = "AnaCoDa Log-Likelihood",
    y = "Custom (Adjusted) Log-Likelihood"
  ) +
  theme_minimal()

```

```{r myll_vs_anacodall_cortest, message=TRUE, warning=TRUE}
# 8. correlation test
cor.test(comparison$AnaCoDa_LL, comparison$LL_sum)
```

# 13. fitting a linear model 
```{r fitting_lm, message=TRUE, warning=TRUE}
lm_fit <- lm(LogLikelihood ~ Gene + Sample, data = loglik_df)

summary(lm_fit)
```

# 14. extract coefficients 
```{r extract_coefs, message=TRUE, warning=TRUE}
coefs <- coef(lm_fit)

#extract gene level coefs 
gene_coefs <- coefs[grep("^Gene", names(coefs))]
```

# get worst 25 genes in the subset 
```{r worst_25, message=TRUE, warning=TRUE}
worst_genes <- sort(gene_coefs)[1:25]
print(worst_genes)
```
# making into a df
```{r worst_df, message=TRUE, warning=TRUE}
worst_df <- enframe(gene_coefs, name = "Gene", value = "Estimate") %>%
  arrange(Estimate) %>%
  slice_head(n = 25)
print(worst_df)
```

#example plot for the 25 worst fitting genes 
```{r worst_plot, message=TRUE, warning=TRUE}
ggplot(worst_df, aes(x = reorder(Gene, Estimate), y = Estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "25 Worst-Fitting Genes by Log-Likelihood",
    x = "Gene",
    y = "Effect on Log-Likelihood"
  ) +
  theme_minimal(base_size = 12)

```

