##SIMULATE: second part of my CUB project
#libraries
```{r}
library(AnaCoDa)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
```

#initialize genome with old genome
```{r}
genome <- initializeGenomeObject(file = "fasta/cleaned_random_1000.fasta")
```

#follow steps in CUB project Rmd to retrieve csp_long and synth_trace from old genome
#initialize parameters
```{r}
parameter <- initializeParameterObject(
  genome = genome, 
  sphi = 1, 
  num.mixtures = 1, 
  gene.assignment = rep(1, length(genome))
)
```

#initialize mcmc
```{r}
mcmc <- initializeMCMCObject(samples = 1000, thinning = 10, adaptive.width = 50)
```

#initialize model 
```{r}
model <- initializeModelObject(parameter = parameter, model = "ROC")
```

#run mcmc
```{r}
runMCMC(mcmc = mcmc, genome = genome, model = model)
```

#save/load results
```{r}
species <- "lachancea_kluyveri"
dir.create("Results", showWarnings = FALSE)

mcmc_file <- file.path("Results", paste0(species, "_mcmc.Rda"))
writeMCMCObject(mcmc = mcmc, file = mcmc_file)

param_file <- file.path("Results", paste0(species, "_parameter.Rda"))
writeParameterObject(parameter = parameter, file = param_file)
```

#extract trace 
```{r}
trace <- parameter$getTraceObject()
plot(x = trace, what = "Expression", mixture = 1, geneIndex = 650)
```

#get synthesis rate trace (φ)
```{r synthesis_trace}
synth_trace_list <- trace$getSynthesisRateTrace()[[1]]

trace_to_tibble <- function(list_trace, id_names) {
  matrix <- do.call(rbind, list_trace)
  sample_n <- 1:ncol(matrix) - 1
  colnames(matrix) <- sample_n
  tibble <- as_tibble(matrix) %>% bind_cols(tibble(id_names), .)
  return(tibble)
}

synth_trace <- trace_to_tibble(synth_trace_list, id_names = getNames(genome))
head(synth_trace)
dim(synth_trace)
str(synth_trace)
```

#get codon-specific parameters
#extract CSP estimates
```{r csp_estimates}
csp_est <- getCSPEstimates(parameter = parameter, mixture = 1, samples = 1000)

csp_codons <- csp_est$Mutation %>% filter(Mean != 0) %>% select("Codon")
str(csp_codons)
print(csp_codons)
```

#extract codon-specific traces
```{r csp_trace}
csp_cat <- c("Mutation", "Selection")
csp_wide_list <- list()

for (i in 1:2) {
  param = csp_cat[i]
  trace_list <- trace$getCodonSpecificParameterTrace((i - 1))[[1]]
  csp_wide_list[[i]] <- trace_to_tibble(trace_list, csp_codons) %>%
    rowwise() %>%
    mutate(AA = codonToAA(Codon), .before = "Codon")
}

names(csp_wide_list) <- csp_cat
```

## pivot traces to long format
```{r csp_long}
csp_long_list <- list()

for (i in 1:2) {
  param = csp_cat[i]
  csp_long_list[[i]] <- pivot_longer(csp_wide_list[[i]],
                                     cols = !c("AA", "Codon"),
                                     values_to = param,
                                     names_to = "Sample") %>%
    mutate(Sample = as.numeric(Sample))
}

names(csp_long_list) <- csp_cat

csp_long <- left_join(csp_long_list$Mutation,
                      csp_long_list$Selection,
                      by = join_by(AA, Codon, Sample))
head(csp_long)
str(csp_long)
```

#then extract phi value
```{r}
phi_sample_650 <- synth_trace %>%
  select(id_names, `650`) %>%
  rename(Mean = `650`)

write.csv(phi_sample_650, "phi_estimates_sample650.csv", row.names = FALSE)
```

#extract mutation (dM) and selection (dEta) values
```{r}
csp_sample_650 <- csp_long %>%
  filter(Sample == 650)

dM_estimates_650 <- csp_sample_650 %>%
  select(Codon, Mutation)

dEta_estimates_650 <- csp_sample_650 %>%
  select(Codon, Selection)

write.csv(dM_estimates_650, "dM_estimates_sample650.csv", row.names = FALSE)
write.csv(dEta_estimates_650, "dEta_estimates_sample650.csv", row.names = FALSE)
```

#checking step
```{r}
# Check if gene names match
phi_ids <- phi_sample_650$id_names
genome_ids <- getNames(genome)

# Check for mismatches
missing_in_phi <- setdiff(genome_ids, phi_ids)
missing_in_genome <- setdiff(phi_ids, genome_ids)

# Print mismatches if any
if (length(missing_in_phi) > 0) {
  cat("Genes missing in phi file:\n")
  print(missing_in_phi)
} else {
  cat("✅ All genome genes are found in phi file!\n")
}

if (length(missing_in_genome) > 0) {
  cat("Genes missing in genome object:\n")
  print(missing_in_genome)
} else {
  cat("✅ All phi genes are found in genome object!\n")
}

```

#simulate
```{r}
simulateGenomeROC <- function(
  genome.obj,
  deta.file,
  dM.file,
  phi.file,
  phi.column = "Mean"
) {
  #' Simulate a genome using parameter estimates from previous model fitting
  #' @param genome.obj: your Genome object (from AnaCoDa)
  #' @param deta.file: CSV file with estimated selection parameters (Δη)
  #' @param dM.file: CSV file with estimated mutation parameters (ΔM)
  #' @param phi.file: CSV file with estimated synthesis rates (ϕ)
  #' @param phi.column: which column to use for phi (default "Mean")
  
  # Load phi estimates
  phi.values <- read.csv(phi.file, header = TRUE)[, phi.column]
  
  # Initialize a new parameter object using your old phi values
  parameter <- initializeParameterObject(
    genome.obj,
    sphi = c(0.01),  # small nonzero sphi (shouldn't matter much for simulation)
    num.mixtures = 1,
    gene.assignment = rep(1, length(genome.obj)),
    mixture.definition = "allUnique",
    initial.expression.values = phi.values,
    split.serine = TRUE
  )
  
  # Load estimated mutation and selection parameters
  parameter$initSelectionCategories(deta.file, 1, FALSE)
  parameter$initMutationCategories(dM.file, 1, FALSE)
  
  # Create a new model object without new phi estimation (fixed phi)
  model <- initializeModelObject(parameter, model = "ROC", with.phi = FALSE)
  
  # Simulate a new genome based on the model parameters
  model$simulateGenome(genome.obj)
  
  return(genome.obj)
}
```

#using sample 650 params
```{r}
# simulate the genome based on your sample 650 parameter estimates
simulated_genome <- simulateGenomeROC(
  genome.obj = genome,
  deta.file = "dEta_estimates_sample650.csv",
  dM.file = "dM_estimates_sample650.csv",
  phi.file = "phi_estimates_sample650.csv",
  phi.column = "Mean"
)
print(simulated_genome)
```

#comparing codon counts 
```{r}
#codon counts from both genomes
subset_codon_counts <- getCodonCounts(genome)
```

```{r}
subset_codon_counts <- getCodonCounts(simulated_genome)
```

#following steps in subset CUB 
#creating samples_eval sequence
```{r}
samples_max <- 1000
samples_eval <- seq(0, samples_max, length.out = 11)
print(samples_eval)
```

#log likelihood loop 
```{r}
#make sure correct LLikGene is loaded**
#getting gene names
gene_names = getNames(simulated_genome)

#creating codon -> amino acid mapping
aa_by_codon <- csp_long %>%
  select(Codon, AA) %>%
  distinct() %>%
  deframe()

loglik_list <- list()
row_index <- 1

# use the filtered codon list for consistency
gene_codons <- csp_codons$Codon
  

for (gene in gene_names) {
  # get codon counts for this gene, using only the modeled codons
  counts <- as.numeric(subset_codon_counts[gene, gene_codons, drop = TRUE])
  names(counts) <- gene_codons
   
  for (s in rev(samples_eval)) {
    # get φ (synthesis rate) for this gene
    phi <- synth_trace %>%
      filter(id_names == gene) %>%
      pull(as.character(s)) %>%
      as.numeric()
    
    # get mutation and selection params
    dM <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Mutation)
    
    dE <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Selection)
    
    # compute log-likelihood for this gene and sample
    ll_result <- LLikGene(gene, dM, dE, phi, counts, aa_by_codon)
    
    # store result
    loglik_list[[row_index]] <- data.frame(
      Gene = gene,
      Sample = s,
      LogLikelihood = ll_result$LogLikelihood
    )
    
    row_index <- row_index + 1
  }
}

# combine results into a tibble for further analysis
loglik_df <- dplyr::bind_rows(loglik_list)
print(loglik_df)
saveRDS(loglik_df, "loglik_df_genome_lk.rds")

loglik_summary <- loglik_df %>%
  group_by(Sample) %>%
  summarise(LL_sum = sum(LogLikelihood), .groups = "drop") 

#save all objects
#saveRDS(loglik_summary, "loglik_summary_lk.rds")

head(loglik_summary)
#summing across samples for one total log-likelihood value per gene
```

#loading both loglik summaries
```{r}
real_loglik <- readRDS("loglik_df_original_genome_lk.rds")
sim_loglik <- readRDS("loglik_df_simulated_genome_lk.rds")
```

#filtering dfs to include specific samples
```{r}
real_filtered <- real_loglik %>% filter(Sample %in% samples_eval)
sim_filtered  <- sim_loglik  %>% filter(Sample %in% samples_eval)

#join by gene and sample 
comparison_loglik <- real_filtered %>%
  rename(Real_LL = LogLikelihood) %>%
  inner_join(sim_filtered %>% rename(Sim_LL = LogLikelihood), 
             by = c("Gene", "Sample")) %>%
  mutate(LL_Difference = Real_LL - Sim_LL)
```


#organize log-likelihoods into a comparison table
```{r}
comparison_loglik <- tibble(
  Gene = getNames(genome),
  Real_LL = real_loglik,
  Sim_LL = sim_loglik,
  LL_Difference = real_loglik - sim_loglik
)

#using sample 1000:
plot_data <- comparison_loglik %>%
  filter(Sample == 1000)

ggplot(plot_data, aes(x = Real_LL, y = Sim_LL)) +
  geom_point(alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Simulated vs Real Log-Likelihood (Sample 1000)",
    x = "Real Log-Likelihood",
    y = "Simulated Log-Likelihood"
  ) +
  theme_minimal()

ggsave("simulated_vs_real_LL.png", width = 8, height = 6, dpi = 300)
```

#quantile calculations
```{r}
real_loglik %>%
  mutate(percent_rank = sum(LogLikelihood>{})/length({}))
```

#extracting the final loglik values
```{r}
obs_df <- real_loglik %>% 
  group_by(Gene) %>% 
  slice_max(order_by = Sample, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  select(Gene, Obs_LL = LogLikelihood)

head(obs_df)
```

#compute empirical quantile
```{r}
quant_df <- sim_loglik %>%                    # many rows per gene
  inner_join(obs_df, by = "Gene") %>%         # bring Obs_LL onto every row
  group_by(Gene) %>%                          # now each group has:
                                              #   • Obs_LL (same for every row)
                                              #   • LogLikelihood (simulated)
  summarise(
    Obs_LL   = first(Obs_LL),                 # pick the single observed value
    Quantile = mean(LogLikelihood <= Obs_LL), # rank / fraction ≤ observed
    .groups  = "drop"
  )

glimpse(quant_df)
```

#plotting quantiles 
```{r}
ggplot(quant_df, aes(x = Quantile)) +
  geom_histogram(
    binwidth = 0.05,              # 20 equal-width bins → easy to eyeball uniformity
    boundary = 0, closed = "left",
    colour = "black", fill = "white"
  ) +
  geom_hline(                      # dashed line = expected count per bin if uniform
    yintercept = nrow(quant_df) / 20,
    linetype = "dashed"
  ) +
  labs(
    title    = "Empirical Quantiles of Observed LLik\nvs. Simulated Distribution",
    x        = "Quantile (0 = lower tail, 1 = upper tail)",
    y        = "Number of genes"
  ) +
  theme_bw()

ggsave("quantile_histogram.png", width = 8, height = 6)
```

