# checking to make sure that my CUB project is calculating everything in the way i think it is
#creating a function that ensures valid cds, a random subset of 500 genes, loads into anacoda
```{r function_random_subset}

library(Biostrings)

create_random_subset_fasta <- function(
  input_fasta,
  output_fasta = "cleaned_random_500.fasta",
  n = 500,
  seed = 42
) {
  # load full fasta
  message("loading fasta...")
  fasta <- readDNAStringSet(input_fasta)

  # filter valid cds (length % 3 == 0 and only a/c/g/t)
  message("filtering valid cds...")
  valid_fasta <- fasta[width(fasta) %% 3 == 0 & !grepl("[^ACGTacgt]", as.character(fasta))]

  # check if there are enough genes to sample
  if (length(valid_fasta) < n) {
    stop("not enough valid genes to sample ", n, " (found only ", length(valid_fasta), ").")
  }

  # randomly sample subset
  set.seed(seed)
  message("sampling ", n, " random genes...")
  subset_indices <- sample(seq_along(valid_fasta), n)
  subset_fasta <- valid_fasta[subset_indices]

  # save to new fasta
  message("saving to: ", output_fasta)
  writeXStringSet(subset_fasta, filepath = output_fasta)

  message("✅ done! file saved: ", output_fasta)
}

create_random_subset_fasta(
  input_fasta = "fasta/revisit_cds_data/candida_tenuis.max.cds",
  output_fasta = "fasta/cleaned_random_500.fasta"
)
```
# 1. load into anacoda 
```{r}
library(AnaCoDa)
genome <- initializeGenomeObject("fasta/cleaned_random_500.fasta")
```

# 2. check if its working
```{r}
{
print(length(genome)) #should be 500
print(str(genome))
}
```

# 3. initialize parameters
```{r}
parameter <- initializeParameterObject(
  genome = genome, 
  sphi = 1, 
  num.mixtures = 1, 
  gene.assignment = rep(1, length(genome))
)
```

# initialize MCMC and model
```{r}
mcmc <- initializeMCMCObject(samples = 100, thinning = 10, adaptive.width = 50)
model <- initializeModelObject(parameter = parameter, model = "ROC")
```

# run MCMC
```{r}
runMCMC(mcmc = mcmc, genome = genome, model = model)
```

# 4. save results 
```{r}
species <- "candida_tenuis"
dir.create("Results", showWarnings = FALSE)

mcmc_file <- file.path("Results", paste0(species, "_mcmc.Rda"))
writeMCMCObject(mcmc = mcmc, file = mcmc_file)

param_file <- file.path("Results", paste0(species, "_parameter.Rda"))
writeParameterObject(parameter = parameter, file = param_file)
```

# 5. trace retrieving and plotting
```{r}
# extract trace object
trace <- parameter$getTraceObject()

# plot expression trace for a single gene (gene 1, mixture 1)
plot(x = trace, what = "Expression", mixture = 1, geneIndex = 1)
```

# 6. retrieving synthesis rate trace 
```{r message=TRUE, warning=TRUE}
synth_trace_list <- trace$getSynthesisRateTrace()[[1]]

trace_to_tibble <- function(list_trace, id_names) {
  matrix <- do.call(rbind, list_trace)
  sample_n <- 1:ncol(matrix) - 1
  colnames(matrix) <- sample_n
  tibble <- as_tibble(matrix) %>% bind_cols(tibble(id_names), .)
  return(tibble)
}

synth_trace <- trace_to_tibble(synth_trace_list, id_names = getNames(genome))
head(synth_trace)
dim(synth_trace)
```

# 7. retrieving codon specific parameters 
```{r csp_estimates, message=TRUE, warning=TRUE}
csp_est <- getCSPEstimates(parameter = parameter, mixture = 1, samples = 100)

csp_codons <- csp_est$Mutation %>% filter(Mean != 0) %>% select("Codon")

str(csp_codons)
print(csp_codons)
```

# extract csp traces
```{r message=TRUE, warning=TRUE}
csp_cat <- c("Mutation", "Selection")
csp_wide_list <- list()

for (i in 1:2) {
  param = csp_cat[i]
  trace_list <- trace$getCodonSpecificParameterTrace((i - 1))[[1]]
  csp_wide_list[[i]] <- trace_to_tibble(trace_list, csp_codons) %>%
    rowwise() %>%
    mutate(AA = codonToAA(Codon), .before = "Codon")
}

names(csp_wide_list) <- csp_cat

```

# pivot traces to long format 
```{r}
csp_long_list <- list()

for (i in 1:2) {
  param = csp_cat[i]
  csp_long_list[[i]] <- pivot_longer(csp_wide_list[[i]],
                                     cols = !c("AA", "Codon"),
                                     values_to = param,
                                     names_to = "Sample") %>%
    mutate(Sample = as.numeric(Sample))
}

names(csp_long_list) <- csp_cat

csp_long <- left_join(csp_long_list$Mutation,
                      csp_long_list$Selection,
                      by = join_by(AA, Codon, Sample))
head(csp_long)
```

#compare between mine and anacoda's
```{r}
#get anacoda’s estimates
csp_estimates <- getCSPEstimates(parameter, mixture = 1, samples = 100)

#combine mutation and selection estimates
csp_summary_ac <- full_join(csp_estimates$Mutation %>% rename(Mutation_mean = Mean),
                            csp_estimates$Selection %>% rename(Selection_mean = Mean),
                            by = c("AA", "Codon"))

#now calculate my means from the trace tibble
csp_summary_tibble <- csp_long %>%
  group_by(AA, Codon) %>%
  summarize(Mutation_calc = mean(Mutation), Selection_calc = mean(Selection), .groups = "drop")

# join the two
csp_comparison <- inner_join(csp_summary_ac, csp_summary_tibble, by = c("AA", "Codon"))

# plot: Mutation
ggplot(csp_comparison, aes(x = Mutation_mean, y = Mutation_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Mutation Parameter: AnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")

# plot: Selection
ggplot(csp_comparison, aes(x = Selection_mean, y = Selection_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Selection Parameter: AnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")
```




