# checking to make sure that my CUB project is calculating everything in the way i think it is
#creating a function that ensures valid cds, a random subset of a number of genes, loads into anacoda
```{r function_random_subset, eval=FALSE, message=TRUE, warning=FALSE}
library(Biostrings)

#function for 500 random
create_random_subset_fasta <- function(
  input_fasta,
  output_fasta = "cleaned_random_500.fasta",
  n = 500,
  seed = 42
) {
  # load full fasta
  message("loading fasta...")
  fasta <- readDNAStringSet(input_fasta)

  # filter valid cds (length % 3 == 0 and only a/c/g/t)
  message("filtering valid cds...")
  valid_fasta <- fasta[width(fasta) %% 3 == 0 & !grepl("[^ACGTacgt]", as.character(fasta))]

  # check if there are enough genes to sample
  if (length(valid_fasta) < n) {
    stop("not enough valid genes to sample ", n, " (found only ", length(valid_fasta), ").")
  }

  # randomly sample subset
  set.seed(seed)
  message("sampling ", n, " random genes...")
  subset_indices <- sample(seq_along(valid_fasta), n)
  subset_fasta <- valid_fasta[subset_indices]

  # save to new fasta
  message("saving to: ", output_fasta)
  writeXStringSet(subset_fasta, filepath = output_fasta)

  message("✅ done! file saved: ", output_fasta)
}

create_random_subset_fasta(
  input_fasta = "fasta/revisit_cds_data/candida_tenuis.max.cds",
  output_fasta = "fasta/cleaned_random_500.fasta"
)
```


```{r function_random_subset, eval=FALSE}
#function for 1000 random 
create_random_subset_fasta <- function(
  input_fasta,
  output_fasta = "cleaned_random_1000.fasta",
  n = 1000,
  seed = 333
) {
  # load full fasta
  message("loading fasta...")
  fasta <- readDNAStringSet(input_fasta)

  # filter valid cds (length % 3 == 0 and only a/c/g/t)
  message("filtering valid cds...")
  valid_fasta <- fasta[width(fasta) %% 3 == 0 & !grepl("[^ACGTacgt]", as.character(fasta))]

  # check if there are enough genes to sample
  if (length(valid_fasta) < n) {
    stop("not enough valid genes to sample ", n, " (found only ", length(valid_fasta), ").")
  }

  # randomly sample subset
  set.seed(seed)
  message("sampling ", n, " random genes...")
  subset_indices <- sample(seq_along(valid_fasta), n)
  subset_fasta <- valid_fasta[subset_indices]

  # save to new fasta
  message("saving to: ", output_fasta)
  writeXStringSet(subset_fasta, filepath = output_fasta)

  message("✅ done! file saved: ", output_fasta)
}

#running lachancea kluyveri
create_random_subset_fasta(
  input_fasta = "fasta/revisit_cds_data/lachancea_kluyveri.max.cds",
  output_fasta = "fasta/cleaned_random_1000.fasta"
)
```

# 1. load into anacoda 
```{r message=TRUE, warning=TRUE}
library(AnaCoDa)
genome <- initializeGenomeObject("fasta/cleaned_random_1000.fasta") #change depending on what subset we want to use
```

#load other necessary libraries
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
```


# 2. check if its working
```{r}
{
print(length(genome)) #should be exact same number of genes in fasta file
print(str(genome))
}
```

# 3. initialize parameters
```{r message=TRUE, warning=TRUE}
parameter <- initializeParameterObject(
  genome = genome, 
  sphi = 1, 
  num.mixtures = 1, 
  gene.assignment = rep(1, length(genome))
)
```

# initialize MCMC and model
```{r message=TRUE, warning=TRUE}
mcmc <- initializeMCMCObject(samples = 5000, thinning = 10, adaptive.width = 50) #4/6 changing samples to 5000
model <- initializeModelObject(parameter = parameter, model = "ROC")
```

# run MCMC
```{r}
runMCMC(mcmc = mcmc, genome = genome, model = model)
```

# 4. save results 
```{r}
species <- "lachancea_kluyveri"
dir.create("Results", showWarnings = FALSE)

mcmc_file <- file.path("Results", paste0(species, "_mcmc_subset_s5000.Rda")) #change name based on samples
writeMCMCObject(mcmc = mcmc, file = mcmc_file)

param_file <- file.path("Results", paste0(species, "_parameter_subset_s5000.Rda")) #change name based on samples
writeParameterObject(parameter = parameter, file = param_file)
```

# 5. trace retrieving and plotting
```{r}
# extract trace object
trace <- parameter$getTraceObject()

# plot expression trace for a single gene (gene 1, mixture 1)
plot(x = trace, what = "Expression", mixture = 1, geneIndex = 200)
```

# 6. retrieving synthesis rate trace 
```{r message=TRUE, warning=TRUE}
synth_trace_list <- trace$getSynthesisRateTrace()[[1]]

trace_to_tibble <- function(list_trace, id_names) {
  matrix <- do.call(rbind, list_trace) #converts list of traces into a matrix
  sample_n <- 1:ncol(matrix) - 1
  colnames(matrix) <- sample_n
  tibble <- as_tibble(matrix) %>% bind_cols(tibble(id_names), .) #create trace tibble
  return(tibble)
}

synth_trace <- trace_to_tibble(synth_trace_list, id_names = getNames(genome))

{
print(head(synth_trace))
print(dim(synth_trace))
}
```

# 7. retrieving codon specific parameters 
```{r codon specific parameters, message=TRUE, warning=TRUE}
csp_est <- getCSPEstimates(parameter = parameter, mixture = 1, samples = 5000, relative.to.optimal.codon = FALSE)

csp_codons <- csp_est$Mutation %>%
  filter(Mean !=0) %>%
  select("Codon")

#codon_vector <- csp_codons$Codon #extracting codon names as a vector

{
print(str(csp_codons))
print(csp_codons)
}
```

# extract csp traces
```{r message=TRUE, warning=TRUE}
csp_cat <- c("Mutation", "Selection")
csp_wide_list <- list()

for(i in 1:2) {
  param = csp_cat[i] ## don't need double [] because already atomic object, mutation for i = 1 and selection for i = 2
  trace_list <- trace$getCodonSpecificParameterTrace((i-1))[[1]]
  csp_wide_list[[i]] <- trace_to_tibble(trace_list, id_names = csp_codons)  %>%
    rowwise() %>%  #use rowwise because codonToAA isn't vectorized.
    mutate(AA = codonToAA(Codon), .before = "Codon")
  
}
names(csp_wide_list) <- csp_cat

csp_long_list <- list()
for(i in 1:2) {
  param = csp_cat[i] ## don't need double [] because already atomic object
  csp_long_list[[i]] <- pivot_longer(csp_wide_list[[i]],
               cols = !c("AA", "Codon"),
               values_to = param,
               names_to = "Sample") %>%
    mutate(Sample = as.numeric(Sample))
}
names(csp_long_list) <- csp_cat

csp_long <- left_join(csp_long_list$Mutation,
                      csp_long_list$Selection,
                      by = join_by(AA, Codon, Sample)
)


{print(head(csp_long))
  print(str(csp_long))
}
```

#compare between mine and anacoda's
```{r}
#get anacoda’s estimates
anacoda_csp_estimates <- getCSPEstimates(parameter, mixture = 1, samples = 5000, relative.to.optimal.codon = FALSE)

#get full 59 codons including reference for comparison sake
full_csp_codons <- csp_est$Mutation %>% select("Codon")
str(full_csp_codons)

#combine mutation and selection estimates
csp_mutation <- anacoda_csp_estimates$Mutation %>% 
  mutate(order = row_number()) %>% 
  dplyr::rename(Mutation_mean = Mean)

csp_selection <- anacoda_csp_estimates$Selection %>% 
  dplyr::rename(Selection_mean = Mean)

#left join using the mutation table as the left side to preserve its order
csp_summary_ac <- left_join(csp_mutation, csp_selection, by = c("AA", "Codon"))

#then, if needed, sort by the order column
csp_summary_ac <- csp_summary_ac %>% arrange(order)

print(head(csp_summary_ac))
print(dim(csp_summary_ac))

#now calculate my means from the trace tibble
csp_summary_tibble <- csp_long %>%
  group_by(AA, Codon) %>%
  summarize(Mutation_calc = mean(Mutation), Selection_calc = mean(Selection), .groups = "drop")

# join the two
csp_comparison <- inner_join(csp_summary_ac, csp_summary_tibble, by = c("AA", "Codon"))
```


```{r}
# plot: mutation
ggplot(csp_comparison, aes(x = Mutation_mean, y = Mutation_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Mutation Parameter: \nAnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")
```

```{r}
# plot: selection
ggplot(csp_comparison, aes(x = Selection_mean, y = Selection_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Selection Parameter: \nAnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")
```

# configuring codons
```{r}
subset_codon_counts <- getCodonCounts(genome)
print(subset_codon_counts)
```

#creating samples_eval sequence
```{r}
samples_max <- 5000
samples_eval <- seq(0, samples_max, length.out = 11)
print(samples_eval)

#for the sake of time, hard coding this 
samples_eval <- c(10, 100, 200, 500, 1000, 5000)

```


# log likelihood loop 
```{r}
#make sure correct LLikGene is loaded**
#getting gene names
gene_names = getNames(genome)

#creating codon -> amino acid mapping
aa_by_codon <- csp_long %>%
  select(Codon, AA) %>%
  distinct() %>%
  deframe()

loglik_list <- list()
row_index <- 1

# use the filtered codon list for consistency
gene_codons <- csp_codons$Codon
  

for (gene in gene_names) {
  # get codon counts for this gene, using only the modeled codons
  counts <- as.numeric(subset_codon_counts[gene, gene_codons, drop = TRUE])
  names(counts) <- gene_codons
   
  for (s in rev(samples_eval)) {
    # get φ (synthesis rate) for this gene
    phi <- synth_trace %>%
      filter(id_names == gene) %>%
      pull(as.character(s)) %>%
      as.numeric()
    
    # get mutation and selection params
    dM <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Mutation)
    
    dE <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Selection)
    
    # compute log-likelihood for this gene and sample
    ll_result <- LLikGene(gene, dM, dE, phi, counts, aa_by_codon)
    
    # store result
    loglik_list[[row_index]] <- data.frame(
      Gene = gene,
      Sample = s,
      LogLikelihood = ll_result$LogLikelihood
    )
    
    row_index <- row_index + 1
  }
}

# combine results into a tibble for further analysis
loglik_df <- dplyr::bind_rows(loglik_list)
print(loglik_df)
saveRDS(loglik_df, "loglik_df_5000s_lachkluv_5.rds")

loglik_summary <- loglik_df %>%
  group_by(Sample) %>%
  summarise(LL_sum = sum(LogLikelihood), .groups = "drop") 
#get rid of grouping to avoid issues

head(loglik_summary)
#summing across samples for one total log-likelihood value per gene
```

#load saved RDS file
```{r}
loglik_df <- readRDS("loglik_df_5000s_lachkluv_5.rds")
```


#plotting overall LL stabilization 
```{r}
ggplot(loglik_summary, aes(x = Sample, y = LL_sum)) +
  geom_line(color = "darkblue") +
  geom_point(color = "darkblue") +
  labs(
    title = "Total Log-Likelihood Over MCMC Samples",
    x = "Sample",
    y = "Total Log-Likelihood"
  ) +
  theme_minimal()
```

# get anacodas llik trace 
```{r}
# extract anacodas log likelihood trace from the MCMC object
logLikeTrace <- mcmc$getLogLikelihoodTrace()
head(logLikeTrace)

acLL_summary <- tibble(
  Sample = 1:length(logLikeTrace[-1]),  # if you want to drop the first (burn-in) sample
  AnaCoDa_LL = logLikeTrace[-1]
)
head(acLL_summary)

#fix constant if necessary
#acLL_summary <- acLL_summary %>%
#  mutate(AnaCoDa_LL_adjusted = AnaCoDa_LL - ###)
#head(acLL_summary)
```

# create separate tibble with just ac value and sample
```{r}
acLL <- acLL_summary %>%
  mutate(Adjusted = AnaCoDa_LL) %>%
  select(Sample, Adjusted)
head(acLL)
```

#plotting my values with ac, need to fix bc not subsetted w same samples 
```{r}
comparison <- left_join(loglik_summary, acLL, by = "Sample")
head(comparison)

comparison_filtered <- comparison %>% filter(Sample != 0)

ggplot(comparison_filtered, aes(x = Adjusted, y = LL_sum)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Custom vs. AnaCoDa Log-Likelihood Comparison",
    x = "Adjusted AnaCoDa LL",
    y = "Custom Total LL"
  ) +
  theme_minimal()
```
#correlation test 
```{r}
cor_test <- cor.test(comparison_filtered$Adjusted, comparison_filtered$LL_sum)
print(cor_test)
```

#plotting deltas 
```{r}
# compute delta values relative to the first valid sample (sample 1)
delta_values <- comparison_filtered %>% 
  arrange(Sample) %>%
  mutate(
    Delta_LL_AnaCoDa = Adjusted - Adjusted[1],
    Delta_LL_Custom  = LL_sum - LL_sum[1]
  )

# check the resulting data
head(delta_values)

# plot the delta values
ggplot(delta_values, aes(x = Delta_LL_AnaCoDa, y = Delta_LL_Custom)) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Delta Log-Likelihood Comparison",
    x = expression(Delta~LogL[AnaCoDa]),
    y = expression(Delta~LogL[Custom])
  ) +
  theme_minimal()
```

#delta cor test
```{r}
delta_cor_test_result <- cor.test(delta_values$Delta_LL_AnaCoDa, delta_values$Delta_LL_Custom, method = "pearson")
print(delta_cor_test_result)

```

#fitting linear model 
```{r}
lm_fit <- lm(LogLikelihood ~ Gene + Sample, data = loglik_df)
summary(lm_fit)
```

#extract coefficients
```{r}
coefs <- coef(lm_fit)
#extract gene level coefs 
gene_coefs <- coefs[grep("^Gene", names(coefs))]
```

#get worst 25 genes 
```{r}
worst_genes <- sort(gene_coefs)[1:25]
print(worst_genes)
```

#making into a df
```{r}
worst_df <- enframe(gene_coefs, name = "Gene", value = "Estimate") %>%
  arrange(Estimate) %>%
  slice_head(n = 25)
print(worst_df)
```

#plotting for the 25 worst fitting genes
```{r}
ggplot(worst_df, aes(x = reorder(Gene, Estimate), y = Estimate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "25 Worst-Fitting Genes by Log-Likelihood",
    x = "Gene",
    y = "Effect on Log-Likelihood"
  ) +
  theme_minimal(base_size = 12)
```

#identifying genes whos llik dropped considerably between a reasonable midpoint and the final point 
```{r}
# determine your midpoint and final sample numbers
mid_sample <- floor(max(loglik_df$Sample) / 2)
final_sample <- max(loglik_df$Sample)

# extract log-likelihood values at the midpoint for each gene
mid_loglik <- loglik_df %>% 
  filter(Sample == mid_sample) %>% 
  select(Gene, LogLikelihood) %>% 
  rename(LogLikelihood_mid = LogLikelihood)

# extract log-likelihood values at the final sample for each gene
final_loglik <- loglik_df %>% 
  filter(Sample == final_sample) %>% 
  select(Gene, LogLikelihood) %>% 
  rename(LogLikelihood_final = LogLikelihood)

# join the two data frames by Gene and compute the drop in log-likelihood
gene_drop <- inner_join(mid_loglik, final_loglik, by = "Gene") %>%
  mutate(LLik_drop = LogLikelihood_mid - LogLikelihood_final)

# flag genes with a drop greater than or equal to the threshold (2 units based on bolker)
threshold <- 2
genes_with_drop <- gene_drop %>% filter(LLik_drop >= threshold)

print(genes_with_drop)
```

#sort by size of drop
```{r}
worst_drops <- gene_drop %>%
  arrange(desc(LLik_drop))

#view top ten worst fitting genes
head(worst_drops, 10)
```

#get top ten worst fitting genes (largest llik drops)
```{r}
top_10_worst <- gene_drop %>%
  arrange(desc(LLik_drop)) %>%
  slice_head(n = 10)
print(top_10_worst)
```

#plot ten worst fitting genes 
```{r}
ggplot(top_10_worst, aes(x = reorder(Gene, LLik_drop), y = LLik_drop)) +
  geom_col(fill = "firebrick") +
  geom_text(aes(label = round(LLik_drop, 2)),
            hjust = -0.1, size = 3.5, color = "black") +
  coord_flip() +
  ylim(0, max(top_10_worst$LLik_drop) + 1.5) +
  labs(
    title = "Top 10 Worst-Fitting Genes by Log-Likelihood Drop",
    x = "Gene",
    y = "LLik Drop (Midpoint − Final)"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title.position = "plot")
```

#plotting histogram
```{r}
gene_drop_hist <- inner_join(mid_loglik, final_loglik, by = "Gene") %>%
  mutate(LogLik_change = LogLikelihood_final - LogLikelihood_mid)

ggplot(gene_drop_hist, aes(x = LogLik_change)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "white") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(
    title = "Change in Log-Likelihood per Gene\n(Final - Midpoint)",
    x = "Log-Likelihood Change",
    y = "Number of Genes"
  ) +
  coord_cartesian(xlim = c(-10,10)) +
  theme_minimal(base_size = 14)
```

#working with larger subset - e.g. 5000s
```{r}
ggplot(loglik_df, aes(x = LogLikelihood)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Sample, scales = "free_x") +
  labs(
    title = "Distribution of Gene Log-Likelihoods at Each Checkpoint",
    x = "Log-Likelihood",
    y = "Number of Genes"
  ) +
  theme_minimal()
```

#calculating drops 
```{r}
# create a tibble with each consecutive sample pair from  samples_eval vector
sample_pairs <- tibble(
  sample_start = samples_eval[-length(samples_eval)],
  sample_end   = samples_eval[-1]
)

# for each pair, compute the log-likelihood drop per gene
loglik_drop_df <- sample_pairs %>%
  rowwise() %>%
  mutate(drop_df = list({
    # get log-likelihoods at the starting checkpoint
    start_df <- loglik_df %>% 
      filter(Sample == sample_start) %>% 
      select(Gene, LogLikelihood) %>%
      rename(LogLikelihood_start = LogLikelihood)
    
    # get log-likelihoods at the ending checkpoint
    end_df <- loglik_df %>% 
      filter(Sample == sample_end) %>% 
      select(Gene, LogLikelihood) %>%
      rename(LogLikelihood_end = LogLikelihood)
    
    # join by gene and compute the drop (start - end)
    inner_join(start_df, end_df, by = "Gene") %>%
      mutate(LLik_drop = LogLikelihood_start - LogLikelihood_end)
  })) %>%
  ungroup() %>%
  unnest(cols = c(drop_df)) %>%
  mutate(pair = paste0(sample_start, "→", sample_end))
```


#plot distributions at specific checkpoints 
```{r}
samples_eval_specific <- c(0, 10, 100, 500, 1000, 2500, 5000)

loglik_eval <- loglik_df %>% filter(Sample %in% samples_eval_specific)

ggplot(loglik_eval, aes(x = LogLikelihood)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Sample, scales = "free_x") +
  labs(
    title = "Gene Log-Likelihood Distributions \nat Selected Samples",
    x = "Log-Likelihood",
    y = "Number of Genes"
  ) +
  theme_minimal()
```

#adding log liks for samples 
```{r}
# run this after loading loglik_df (if you want to append to it)
samples_to_add <- c(10, 100, 200)
loglik_list_new <- list()
row_index <- 1

for (gene in gene_names) {
  counts <- as.numeric(subset_codon_counts[gene, gene_codons, drop = TRUE])
  names(counts) <- gene_codons

  for (s in samples_to_add) {
    phi <- synth_trace %>%
      filter(id_names == gene) %>%
      pull(as.character(s)) %>%
      as.numeric()

    dM <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Mutation)

    dE <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Selection)

    ll_result <- LLikGene(gene, dM, dE, phi, counts, aa_by_codon)

    loglik_list_new[[row_index]] <- data.frame(
      Gene = gene,
      Sample = s,
      LogLikelihood = ll_result$LogLikelihood
    )

    row_index <- row_index + 1
  }
}

# create the new entries
new_entries <- bind_rows(loglik_list_new)

# only add truly new rows (avoid duplicate Gene+Sample)
loglik_df <- bind_rows(
  loglik_df,
  anti_join(new_entries, loglik_df, by = c("Gene", "Sample"))
)

# optional: save the updated loglik_df
saveRDS(loglik_df, "loglik_df_CLEAN.rds")

str(loglik_df)
unique(loglik_df$Sample)
```


#FOR PRESENTATION. START HERE.
#looking at new loglikdf
```{r}
library(stringr)
loglik_df <- readRDS("loglik_df_5000s_lachkluv_5.rds") %>%
  mutate(
    chrom = str_extract(Gene, "SAKL0[A-Z]"),     # pulls out "SAKL0C"
    chrom = str_extract(chrom, "[A-Z]$")         # pulls out just "C"
  ) %>%
  filter(Sample != 0) 

str(loglik_df)
names(loglik_df)
head(loglik_df)
print(loglik_df)
```

#calculate and plot
```{r}
new_pairs <- tibble(
  sample_start = c(100, 200, 500, 1000),
  sample_end = c(5000, 5000, 5000, 5000)
)

new_drop_df <- new_pairs %>%
  rowwise() %>%
  mutate(drop_df = list({
    start_df <- loglik_df %>%
      filter(Sample == sample_start) %>%
      select(Gene, LogLikelihood, chrom) %>%
      rename(LogLikelihood_start = LogLikelihood)

    end_df <- loglik_df %>%
      filter(Sample == sample_end) %>%
      select(Gene, LogLikelihood) %>%
      rename(LogLikelihood_end = LogLikelihood)

    inner_join(start_df, end_df, by = "Gene") %>%
      mutate(LLik_drop = LogLikelihood_end - LogLikelihood_start)
  })) %>%
  ungroup() %>%
  unnest(cols = c(drop_df)) %>%
  mutate(pair = paste0(sample_start, "→", sample_end))

str(new_drop_df)
unique(new_drop_df$pair)
```

#replot histogram
```{r}
histograms <- ggplot(new_drop_df, aes(x = LLik_drop)) +
  geom_histogram(bins = 80, fill = "mediumpurple", color = "white") +
  geom_vline(xintercept = -2, linetype = "dashed", color = "red", size = 0.3) +
  facet_wrap(~ sample_start) +
  coord_cartesian(xlim = c(-10, 10)) +
  scale_y_log10() +
  labs(
    title = "Distribution of Log-Likelihood Drops \nBetween Checkpoints",
    x = "Log-Likelihood Drop (negative = worsening fit, end - start)",
    y = "Number of Genes (Log Scale)"
  ) +
  theme_minimal(base_size = 14) +  # global font size base
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13),
    strip.text = element_text(size = 14, face = "bold")  # facet labels
  )

plot(histograms)
#save it
ggsave("loglik_drop_histogram.png", histograms, width = 10, height = 10, dpi = 300, bg = "white")
```

#200 -> 5000 focus
```{r}
drop_200_df <- new_drop_df %>%
  filter(sample_start == 200, sample_end == 5000)

str(drop_200_df)
drop_200_df <- drop_200_df %>%
  mutate(outlier = LLik_drop > 0)

pr_drop_by_chrom <- drop_200_df %>%
  group_by(chrom) %>%
  summarise(
    total_genes = n(),
    num_drops = sum(outlier),
    PrDrop = num_drops / total_genes
  ) %>%
  mutate(highlight = chrom == "C")

proofforchromc <- ggplot(pr_drop_by_chrom, aes(x = reorder(chrom, -PrDrop), y = PrDrop, fill = highlight)) +
  geom_col(color = "black", width = 0.75) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "gray80")) +
  labs(
    title = "Proportion of Genes with LLik Drop \n> 2 by Chromosome (200 → 5000)",
    subtitle = "Chromosome C highlighted in red",
    x = "Chromosome",
    y = "P(Drop > 2)",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12)
  )

ggsave("proof_chromosome_c.png", plot = proofforchromc, width = 8, height = 6, dpi = 300, bg = "white")

```

#creating overlapping chrom figure
```{r}
overlappingchrom <- ggplot(drop_200_df, aes(x = LLik_drop, color = chrom)) +
  geom_density(alpha = 0.3) +
  coord_cartesian(xlim = c(-2, 10)) +
  labs(
    title = "Δ Log-Likelihoods (200 → 5000) \nby Chromosome",
    x = "LLik Drop",
    y = "Density"
  ) +
  theme_minimal(base_size = 13)

ggsave("overlappingchrom.png", plot = overlappingchrom, width = 8, height = 6, dpi = 300, bg = "white")

```

#plot all genes showing full distrib of llik_drop in drop 200 df
```{r}
fulldistrib200 <- ggplot(drop_200_df, aes(x = LLik_drop)) +
  # gray rectangle for LLik_drop > 0
  annotate("rect", xmin = 0, xmax = Inf, ymin = -Inf, ymax = Inf, 
           fill = "gray90", alpha = 0.5) +
  
  # histogram
  geom_histogram(bins = 60, fill = "steelblue", color = "white") +
  
  # red dashed line at threshold
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 0.6) +
  
  # zoom in on relevant range
  coord_cartesian(xlim = c(-5, 10)) +
  
  labs(
    title = "Distribution of Δ Log-Likelihoods (200 → 5000)",
    subtitle = "Red dashed line = deviance threshold (Δ < 0)\nShaded region = improved fit (Δ > 0)",
    x = "LLik Drop",
    y = "Number of Genes"
  ) +
  
  theme_minimal(base_size = 14)

ggsave("fulldistrib200.png", plot = fulldistrib200, width = 8, height = 6, dpi = 300, bg = "white")
```

#zoomed in plot 
```{r}
drop_200_df <- drop_200_df %>%
  mutate(outlier = LLik_drop > 0)

pr_drop_neg_by_chrom <- drop_200_df %>%
  group_by(chrom) %>%
  summarise(
    total_genes = n(),
    num_drops = sum(outlier),
    PrDrop = num_drops / total_genes
  ) %>%
  mutate(highlight = chrom == "C")

deviant_plot <- ggplot(pr_drop_neg_by_chrom, aes(x = reorder(chrom, -PrDrop), y = PrDrop, fill = highlight)) +
  geom_col(color = "black", width = 0.75) +
  scale_fill_manual(values = c("TRUE" = "firebrick", "FALSE" = "gray80")) +
  labs(
    title = "Proportion of Genes with LLik Drop < 0\n by Chromosome (200 → 5000)",
    subtitle = "Chromosome C highlighted in red",
    x = "Chromosome",
    y = "P(Drop < 0)",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.text = element_text(size = 12),
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 12)
  )

ggsave("zoomdeviantplot.png", plot = deviant_plot, width = 8, height = 6, dpi = 300, bg = "white")
```


