# checking to make sure that my CUB project is calculating everything in the way i think it is
#creating a function that ensures valid cds, a random subset of 500 genes, loads into anacoda
```{r function_random_subset, eval=FALSE}

library(Biostrings)

create_random_subset_fasta <- function(
  input_fasta,
  output_fasta = "cleaned_random_500.fasta",
  n = 500,
  seed = 42
) {
  # load full fasta
  message("loading fasta...")
  fasta <- readDNAStringSet(input_fasta)

  # filter valid cds (length % 3 == 0 and only a/c/g/t)
  message("filtering valid cds...")
  valid_fasta <- fasta[width(fasta) %% 3 == 0 & !grepl("[^ACGTacgt]", as.character(fasta))]

  # check if there are enough genes to sample
  if (length(valid_fasta) < n) {
    stop("not enough valid genes to sample ", n, " (found only ", length(valid_fasta), ").")
  }

  # randomly sample subset
  set.seed(seed)
  message("sampling ", n, " random genes...")
  subset_indices <- sample(seq_along(valid_fasta), n)
  subset_fasta <- valid_fasta[subset_indices]

  # save to new fasta
  message("saving to: ", output_fasta)
  writeXStringSet(subset_fasta, filepath = output_fasta)

  message("✅ done! file saved: ", output_fasta)
}

create_random_subset_fasta(
  input_fasta = "fasta/revisit_cds_data/candida_tenuis.max.cds",
  output_fasta = "fasta/cleaned_random_500.fasta"
)
```
# 1. load into anacoda 
```{r message=TRUE, warning=TRUE}
library(AnaCoDa)
genome <- initializeGenomeObject("fasta/cleaned_random_500.fasta")
```

# 2. check if its working
```{r}
{
print(length(genome)) #should be 500
print(str(genome))
}
```

# 3. initialize parameters
```{r message=TRUE, warning=TRUE}
parameter <- initializeParameterObject(
  genome = genome, 
  sphi = 1, 
  num.mixtures = 1, 
  gene.assignment = rep(1, length(genome))
)
```

# initialize MCMC and model
```{r message=TRUE, warning=TRUE}
mcmc <- initializeMCMCObject(samples = 100, thinning = 10, adaptive.width = 50)
model <- initializeModelObject(parameter = parameter, model = "ROC")
```

# run MCMC
```{r}
runMCMC(mcmc = mcmc, genome = genome, model = model)
```

# 4. save results 
```{r}
species <- "candida_tenuis"
dir.create("Results", showWarnings = FALSE)

mcmc_file <- file.path("Results", paste0(species, "_mcmc_subset.Rda"))
writeMCMCObject(mcmc = mcmc, file = mcmc_file)

param_file <- file.path("Results", paste0(species, "_parameter_subset.Rda"))
writeParameterObject(parameter = parameter, file = param_file)
```

# 5. trace retrieving and plotting
```{r}
# extract trace object
trace <- parameter$getTraceObject()

# plot expression trace for a single gene (gene 1, mixture 1)
plot(x = trace, what = "Expression", mixture = 1, geneIndex = 1)
```

# 6. retrieving synthesis rate trace 
```{r message=TRUE, warning=TRUE}
synth_trace_list <- trace$getSynthesisRateTrace()[[1]]

trace_to_tibble <- function(list_trace, id_names) {
  matrix <- do.call(rbind, list_trace) #converts list of traces into a matrix
  sample_n <- 1:ncol(matrix) - 1
  colnames(matrix) <- sample_n
  tibble <- as_tibble(matrix) %>% bind_cols(tibble(id_names), .) #create trace tibble
  return(tibble)
}

synth_trace <- trace_to_tibble(synth_trace_list, id_names = getNames(genome))

{
print(head(synth_trace))
print(dim(synth_trace))
}
```

# 7. retrieving codon specific parameters 
```{r codon specific parameters, message=TRUE, warning=TRUE}
csp_est <- getCSPEstimates(parameter = parameter, mixture = 1, samples = 100, relative.to.optimal.codon = FALSE)

csp_codons <- csp_est$Mutation %>%
  filter(Mean !=0) %>%
  select("Codon")

codon_vector <- csp_codons$Codon #extracting codon names as a vector

{
print(str(csp_codons))
print(csp_codons)
}
```

# extract csp traces
```{r message=TRUE, warning=TRUE}
csp_cat <- c("Mutation", "Selection")
csp_wide_list <- list()

for(i in 1:2) {
  param = csp_cat[i] ## don't need double [] because already atomic object, mutation for i = 1 and selection for i = 2
  trace_list <- trace$getCodonSpecificParameterTrace((i-1))[[1]]
  csp_wide_list[[i]] <- trace_to_tibble(trace_list, id_names = csp_codons)  %>%
    rowwise() %>%  #use rowwise because codonToAA isn't vectorized.
    mutate(AA = codonToAA(Codon), .before = "Codon")
  
}
names(csp_wide_list) <- csp_cat

csp_long_list <- list()
for(i in 1:2) {
  param = csp_cat[i] ## don't need double [] because already atomic object
  csp_long_list[[i]] <- pivot_longer(csp_wide_list[[i]],
               cols = !c("AA", "Codon"),
               values_to = param,
               names_to = "Sample") %>%
    mutate(Sample = as.numeric(Sample))
}
names(csp_long_list) <- csp_cat

csp_long <- left_join(csp_long_list$Mutation,
                      csp_long_list$Selection,
                      by = join_by(AA, Codon, Sample)
)


{print(head(csp_long))
  print(str(csp_long))
}
```

#compare between mine and anacoda's
```{r}
#get anacoda’s estimates
anacoda_csp_estimates <- getCSPEstimates(parameter, mixture = 1, samples = 100)

#get full 59 codons including reference for comparison sake
full_csp_codons <- csp_est$Mutation %>% select("Codon")
str(full_csp_codons)

#combine mutation and selection estimates
csp_mutation <- anacoda_csp_estimates$Mutation %>% 
  mutate(order = row_number()) %>% 
  dplyr::rename(Mutation_mean = Mean)

csp_selection <- anacoda_csp_estimates$Selection %>% 
  dplyr::rename(Selection_mean = Mean)

#left join using the mutation table as the left side to preserve its order
csp_summary_ac <- left_join(csp_mutation, csp_selection, by = c("AA", "Codon"))

#then, if needed, sort by the order column
csp_summary_ac <- csp_summary_ac %>% arrange(order)

print(head(csp_summary_ac))
print(dim(csp_summary_ac))

#now calculate my means from the trace tibble
csp_summary_tibble <- csp_long %>%
  group_by(AA, Codon) %>%
  summarize(Mutation_calc = mean(Mutation), Selection_calc = mean(Selection), .groups = "drop")

# join the two
csp_comparison <- inner_join(csp_summary_ac, csp_summary_tibble, by = c("AA", "Codon"))
```


```{r}
# plot: mutation
ggplot(csp_comparison, aes(x = Mutation_mean, y = Mutation_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Mutation Parameter: AnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")
```

```{r}
# plot: selection
ggplot(csp_comparison, aes(x = Selection_mean, y = Selection_calc)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "blue") +
  labs(title = "Selection Parameter: AnaCoDa vs. Calculated from Trace",
       x = "AnaCoDa Mean", y = "Tibble Mean")

```

# configuring codons
```{r}
subset_codon_counts <- getCodonCounts(genome)
print(subset_codon_counts)
```

# log likelihood loop 
```{r}
#getting gene names
gene_names = getNames(genome)

#creating codon -> amino acid mapping
aa_by_codon <- csp_long %>%
  select(Codon, AA) %>%
  distinct() %>%
  deframe()

loglik_list <- list()
row_index <- 1

# Use the filtered codon list for consistency
gene_codons <- csp_codons$Codon

for (gene in gene_names) {
  # get codon counts for this gene, using only the modeled codons
  counts <- as.numeric(codon_counts[gene, gene_codons, drop = TRUE])
  names(counts) <- gene_codons
  
  for (s in 0:100) {
    # get φ (synthesis rate) for this gene and sample
    phi <- synth_trace %>%
      filter(id_names == gene) %>%
      pull(as.character(s)) %>%
      as.numeric()
    
    # get mutation and selection params for the modeled codons at this sample
    dM <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Mutation)
    
    dE <- csp_long %>%
      filter(Codon %in% gene_codons, Sample == s) %>%
      arrange(match(Codon, gene_codons)) %>%
      pull(Selection)
    
    # compute log-likelihood for this gene and sample
    ll_result <- LLikGene(gene, dM, dE, phi, counts, aa_by_codon)
    
    # store result
    loglik_list[[row_index]] <- data.frame(
      Gene = gene,
      Sample = s,
      LogLikelihood = ll_result$LogLikelihood
    )
    
    row_index <- row_index + 1
  }
}

# combine results into a tibble for further analysis
loglik_df <- dplyr::bind_rows(loglik_list)
print(loglik_df)


loglik_summary <- loglik_df %>%
  group_by(Sample) %>%
  summarise(LL_sum = sum(LogLikelihood), .groups = "drop") 
#get rid of grouping to avoid issues

head(loglik_summary)
#summing across samples for one total log-likelihood value per gene
```

# get anacodas llik trace 
```{r}
# extract anacodas log likelihood trace from the MCMC object
logLikeTrace <- mcmc$getLogLikelihoodTrace()
head(logLikeTrace)

acLL_summary <- tibble(
  Sample = 1:length(logLikeTrace[-1]),  # if you want to drop the first (burn-in) sample
  AnaCoDa_LL = logLikeTrace[-1]
)
head(acLL_summary)

#subtract constant of 267,100
acLL_summary <- acLL_summary %>%
  mutate(AnaCoDa_LL_adjusted = AnaCoDa_LL - 267100)
head(acLL_summary)

```

# create separate tibble with just adjusted value and sample
```{r}
acLL_adjusted <- acLL_summary %>%
  mutate(Adjusted = AnaCoDa_LL - 267100) %>%
  select(Sample, Adjusted)
head(acLL_adjusted)
```

#plotting my values with ac
```{r}
comparison <- left_join(loglik_summary, acLL_adjusted, by = "Sample")
head(comparison)

comparison_filtered <- comparison %>% filter(Sample != 0)

ggplot(comparison_filtered, aes(x = Adjusted, y = LL_sum)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Custom vs. Adjusted AnaCoDa Log-Likelihood Comparison",
    x = "Adjusted AnaCoDa LL",
    y = "Custom Total LL"
  ) +
  theme_minimal()



```
#correlation test 
```{r}
cor_test <- cor.test(comparison_filtered$Adjusted, comparison_filtered$LL_sum)
print(cor_test)
```

